On Solving the Problem of Finding Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction Using~a Parallel Global Search Algorithm

### 2022-10-11 2022/PAVT Special Issue/mathematics-1918483 -for proof.tex

%=================================================================
\documentclass[mathematics,article,submit,pdftex,moreauthors]{Definitions/mdpi} 
% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal and change "submit" to "accept". The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

@@

%=================================================================
% MDPI internal commands
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2022}
\copyrightyear{2022}
%\externaleditor{Academic Editor: Firstname Lastname}
\datereceived{} 
%\daterevised{} % Only for the journal Acoustics
\dateaccepted{} 
\datepublished{} 
%\datecorrected{} % Corrected papers include a "Corrected: XXX" date in the original paper.
%\dateretracted{} % Corrected papers include a "Retracted: XXX" date in the original paper.
\hreflink{https://doi.org/} % If needed use \linebreak
%\doinum{}
%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{On Solving the Problem of Finding Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction Using Parallel Global Search Algorithm}
% Дежурное название: Использование ...алгоритма для определения кинетических параметров процесса каталитической изомеризации пентан-гексановой фракции
% Еще один вариант: О решении задачи определения Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction с помощью Parallel Global Search Algorithm
% На английском: On Solving the Problem of Finding Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction Using Parallel Global Search Algorithm

% MDPI internal command: Title for citation in the left column
\TitleCitation{On Solving the Problem of Finding Kinetic Parameters}

@@

% Authors, for the paper (add full first names)
\Author{Konstantin Barkalov $^{1}$\orcidA{}, Irek Gubaydullin $^{2,3}$, Evgeny Kozinov$^{1}$\orcidB{}, Ilya Lebedev $^{1}$\orcidC{}, Roza Faskhutdinova\orcidD{} $^{2,3}$, Azamat Faskhutdinov$^{2}$ and Leniza Enikeeva $^{3,4,}$*\orcidE{}}

@@

\address{%
$^{1}$ \quad Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, Russia; konstantin.barkalov@itmm.unn.ru\\
$^{2}$ \quad Institute of Petrochemistry and Catalysis – Subdivision of the Ufa Federal Research Centre of RAS, Ufa, Russia; irekmars@mail.ru \\
$^{3}$ \quad Ufa State Petroleum Technological University, Ufa, Russia; roza\_fask@mail.ru\\
$^{4}$ \quad Novosibirsk State University, Novosibirsk, Russia; leniza.enikeeva@yandex.ru
}

% Contact information of the corresponding author
\corres{Correspondence: leniza.enikeeva@yandex.ru (E.L.); konstantin.barkalov@itmm.unn.ru (B.K.) }

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{The article is devoted to the problem of developing a kinetic model of a complex chemical reaction using a parallel optimization method. The design of the kinetic model consists in finding the kinetic parameters of the reaction, which cannot be calculated analytically, and since the chemical reaction involves many stages, the optimization problem is multi-extreme. As a chemical reaction, the process of catalytic isomerization of the pentane-hexane fraction is considered, which is now important due to the switch of the oil refining industry to the production of gasoline corresponding to the Euro-5 standard. To solve the optimization problem, the authors have developed parallel global search algorithm and a program based on Lipschitz optimization. The kinetic parameters found made it possible to develop a mathematical model of the process, which gives good agreement with industrial data.}

% Keywords
\keyword{kinetic model; inverse problems of chemical kinetics; mathematical modeling; catalytic isomerisation of pentane-hexane fraction; Lipschitz optimization; black-box functions; parallel computing}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
%\MSC{}
%\JEL{}

@@

Due to the transition of the domestic oil refining industry to the production of gasoline conforming to the Euro-5 standard, an urgent task is to reduce the content of aromatic hydrocarbons and, in particular, benzene in motor fuel while maintaining the octane value. In connection with the transition of the oil refining industry to the production of motor gasoline conforming to Euro-5 and Euro-6 standards, the reduction of aromatic hydrocarbons in motor fuel and, mainly, benzene becomes an urgent task. At the same time, removing aromatic hydrocarbons, which are high-octane components, it is necessary to preserve the octane number of gasoline. It is known that hydrocarbons of branched structure have a higher octane number than hydrocarbons of linear structure. For example, in $n$-pentane, the octane number according to the research method is 62 points, and in $iso$-pentane it is 93 points. On an industrial scale, catalytic isomerization units of light paraffins have been introduced to produce hydrocarbons of an isomeric structure. The catalytic isomerization of light paraffins, in turn, makes it possible to obtain a high-octane component of automobile gasoline with a minimum content of aromatic hydrocarbons. The main advantage of this process is the production of high-octane gasoline, while aromatic hydrocarbons are practically not formed. Straight-run gasoline fractions with boiling limits of 62-70 $^0$C containing mainly $n$-pentane and $n$-hexane, pentane-hexane fractions from gas condensate or from gas fractionation plants, as well as catalytic reforming raffinates are used as feedstock for catalytic isomerization. Modern bifunctional catalysts are used to carry out the process. In order to prevent coke deposition, the catalytic isomerization process is carried out in a hydrogen-containing gas environment under pressure.

Computer information technologies are widely used in the modeling of production processes and today it is difficult to imagine research activities without the use of modern means of constructing and using models. Moreover, special attention is currently being paid to the development of a detailed kinetic model of complex industrial processes. Based on these models, it is possible to significantly increase the efficiency of the process, while minimally spending energy and material resources. The main advantage of such methods is the absence of additional costs for the construction of pilot plants. The developed model should describe the process within a wide range of operating conditions, have high flexibility, allowing for changes, calculations of the composition of the reaction mixture, etc.

In the article \cite{enik2021} we have already reviewed this process, but in order to simplify the model, the scheme of chemical transformations, in particular hydrocracking reactions, have been revised. Thus, the most probable hydrocracking reactions based on the carbonium-ion mechanism were formed. The new transformation scheme implies the recalculation of the kinetic parameters of the reaction, which form a mathematical model of the process.

In the article \cite{enik2021}, a genetic algorithm was used to calculate kinetic parameters, and the calculation time of kinetic parameters was about 10 hours. Therefore, in order to develop a mathematical model of a new transformation scheme, the question of choosing a more effective optimization method arises.

In the complex problems of chemical kinetics, kinetic constants of reactions cannot be found analytically, they can only be found using numerical methods (see, for example, \cite{Zaynullin2020,Enikeeva2020,Koledina2019}). In this case, the quality criterion (objective function) is set in the form of a certain algorithm, which includes the stage of numerical modeling, which, in turn, makes it compute-intensive. Furthermore, in the inverse problems of chemical kinetics, the objective function can be substantially multiextremal, i.e., having many local extremes along the global one.

Computational algorithms for solving such multiextremal problems (global optimization methods) differ vastly from local optimization techniques since they require the study of the entire search domain (see, for example, \cite{PaulaviciusZilinskas2014,Sergeyev2017}). These methods can be divided into two classes: metaheuristic and deterministic. Metaheuristic algorithms are usually based on simulations of processes occurring in the wild.
Examples of metaheuristic algorithms are simulated annealing, evolutionary computing, etc. (see, for example, \cite{Battiti2009,Eiben2015}).
These algorithms are included in many software libraries for solving global optimization problems (for example, MATLAB Global Optimization Toolbox, Scipy.Optimize, etc.).
Due to the operation speed and low requirements for computing resources (in particular, memory), metaheuristic algorithms have become widespread for solving practical problems. However, the solution to the problem found by such an algorithm is, in general, local and can be located far from the global optimum \cite{Kvasov2018}.

It is also possible to construct deterministic methods of global search, which are different from brute force algorithms and metaheuristic algorithms. Such methods are associated with the presence and consideration of a priori assumptions about the problem properties. 
One of the natural assumptions about the problem is that changing the values of an objective function is limited by changes of its argument. In this case, the function satisfies the Lipschitz condition, and the problem itself is called the Lipschitz global optimization problem. The central direction of the development of modern methods of Lipschitz optimization is the development of parallel algorithms. It is assumed that the values of the objective function are determined as a result of solving some other task and require a costly computational experiment.

This paper reflects the results of applying parallel methods of Lipschitz optimization developed in \cite{Barkalov2016,Strongin2018,Sysoyev2017}, to the solution of inverse problems of chemical kinetics.  
The main part of the paper is structured as follows.

Section \ref{sec_matmodel} provides a description of the mathematical model of the chemical reaction under study. Section \ref{sec_GSA} discusses the standard formulation of the Lipschitz global optimization problem and the algorithm for solving it. Section \ref{sec_numexper} outlines the issues related to the technical aspects of the implementation of the parallel algorithm. The physico-chemical interpretation of the results of numerical calculations using the developed parallel algorithm is presented in the Section \ref{sec_discussion}. 
%The last section contains the conclusion and plans for future work.

\section{Mathematical model}\label{sec_matmodel}

Let's consider the mechanism of chemical transformations of the isomerization of the pentane-hexane fraction. The reaction consists of 48 reaction stages, which are presented in the table \ref{tab_chemistry}. The following abbreviations were adopted: 2-MP is 2-methylpentane; 3-MP is 3-methylpentane; 2,2-DMB is 2,2-dimethylbutane; 2,3-DMB is 2,3-dimethylbutane; CH* is cyclohexane; MCP is methyl cyclopentane; B is benzene; CP is cyclopentane.

\begin{table}[H] 
\caption{Chemical transformations of the catalytic isomerization of the pentane-hexane fraction.\label{tab_chemistry}}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\textwidth}{CC}
\toprule
 \textbf{Reaction stages} & \textbf{Reaction stages} \\
\midrule
1) $n$-C$_5$H$_{12}$ $\rightarrow$ $iso$-C$_5$H$_{12}$ & 25) $n$-C$_6$H$_{14}$ $\rightarrow$ CH* + H$_2$\\
2) $iso$-C$_5$H$_{12}$ $\rightarrow$ $n$-C$_5$H$_{12}$ & 26) CH* + H$_2$ $\rightarrow$ $n$-C$_6$H$_{14}$ \\
3) $n$-C$_6$H$_{14}$ $\rightarrow$ 2-MP & 27) 2-MP $\rightarrow$ MCP+ H$_2$ \\
4) 2-MP $\rightarrow$ $n$-C$_6$H$_{14}$ & 28) MCP + H$_2$ $\rightarrow$ 2-MP  \\
5) $n$-C$_6$H$_{14}$ $\rightarrow$ 3-MP & 29) 3-MP $\rightarrow$ MCP + H$_2$ \\
6) 3-MP $\rightarrow$ $n$-C$_6$H$_{14}$  & 30) MCP + H$_2$ $\rightarrow$ 3-MP \\
7)  2-MP $\rightarrow$ 3-MP & 31) 2,2-DMB $\rightarrow$ MCP + H$_2$ \\
8) 3-MP $\rightarrow$ 2-MP  & 32) MCP + H$_2$ $\rightarrow$ 2,2-DMB \\
9) 2-MP $\rightarrow$ 2,2-DMB  & 33) 2,3-DMB $\rightarrow$ MCP + H$_2$ \\
10) 2,2-DMB $\rightarrow$ 2-MP  & 34) MCP + H$_2$ $\rightarrow$ 2,3-DMB \\
11) 2-MP $\rightarrow$ 2,3-DMB  & 35) CP + H$_2$ $\rightarrow$ $n$-C$_5$H$_{12}$ \\
12) 2,3-DMB $\rightarrow$ 2-MP  & 36) $n$-C$_5$H$_{12}$ $\rightarrow$ CP + H$_2$ \\
13) 3-MP $\rightarrow$ 2,2-DMB  & 37) $n$-C$_4$H$_{10}$ $\rightarrow$ $iso$-C$_4$H$_{10}$ \\
14) 2,2-DMB $\rightarrow$ 3-MP  & 38) $iso$-C$_4$H$_{10}$ $\rightarrow$ $n$-C$_4$H$_{10}$ \\
15) 3-MP $\rightarrow$ 2,3-DMB  & 39) $n$-C$_5$H$_{12}$ + H$_2$ $\rightarrow$ C$_3$H$_8$ + C$_2$H$_6$ \\
16) 2,3-DMB $\rightarrow$ 3-MP  & 40) $n$-C$_5$H$_{12}$ + H$_2$ $\rightarrow$ $n$-C$_4$H$_{10}$ + CH$_4$ \\
17) 2,2-DMB $\rightarrow$ 2,3-DMB  & 41) $iso$-C$_5$H$_{12}$ + H$_2$ $\rightarrow$ $iso$-C$_4$H$_{10}$ + CH$_4$ \\
18) 2,3-DMB $\rightarrow$ 2,2-DMB & 42) $n$-C$_6$H$_{14}$ + H$_2$ $\rightarrow$ 2C$_3$H$_8$ \\
19) B + 3H$_2$ $\rightarrow$ CH*  & 43) $n$-C$_6$H$_{14}$ + H$_2$ $\rightarrow$ $n$-C$_5$H$_{12}$ + CH$_4$ \\
20) CH* $\rightarrow$ B + 3H$_2$ & 44) $n$-C$_6$H$_{14}$ + H$_2$ $\rightarrow$ $n$-C$_4$H$_{10}$ + C$_2$H$_6$ \\
21) B + 3H$_2$ $\rightarrow$ MCP  & 45) 2-MP + H$_2$ $\rightarrow$ $iso$-C$4$H$_{10}$ + C$_2$H$_6$ \\
22) MCP $\rightarrow$ B + 3H$_2$  & 46) 3-MP + H$_2$ $\rightarrow$ $iso$-C$_5$H$_{12}$ + CH$_4$ \\
23) CH* $\rightarrow$ MCP & 47) 2,2-DMB + H$_2$ $\rightarrow$ $iso$-C$_5$H$_{12}$ + CH$_4$ \\
24) MCP $\rightarrow$ CH*  & 48) 2,3-DMB + H$_2$ $\rightarrow$ $iso$-C$_5$H$_{12}$ + CH$_4$ \\
\bottomrule
\end{tabularx}
\end{table}

Mathematical model of the chemical process is a non-linear system of differential equations \cite{Koledina2016KineticsAM,enik2021,logan}:
\begin{linenomath}
\begin{equation}\label{eq_concentrations}
\dfrac{dx_i}{d\tau} = \sum_{j=1}^J \nu_{ij} w_j, i = 1,...,I,
\end{equation}
\begin{equation}\label{eq_rate}
w_{j} = k_j \prod_{i=1}^I \left( \dfrac{x_i}{F} \right),
\end{equation}
\begin{equation}\label{eq_constant}
k_j = k^0_j \exp \left( -\dfrac{E_j}{RT} \right)
\end{equation}
\end{linenomath}
with initial values: at $\tau = 0$, $x_i(0) = x_i^0$, where $\nu_{ij}$ are stoichiometric coefficients of stages of chemical reaction; $J$ is number of stages ($J = 48$), $x_i$ is molar flow rate of the $i$-th component of reaction, kmol/h; $I$ is the number of components ($I = 17$); $w_j$ is the rate of the $j$-th stage of reaction, kmol/(h$\cdot$kg cat.), $E_j$ is activation energy of reaction stage, J/mol; $R$ is the universal gas constant ($R$=8,31 J/(mol$\cdot$K), $T$ is temperature, K; $k_j$ are pre-exponentional factors, kmol/(h$\cdot$kg cat.); $\tau$ is conditioned contact time, (kg cat.); $F$ is the total molar flow rate, kmol/h.

The mathematical model involves such parameters as the raw material flow rate $F$ in the reactor and temperature $T$, but these two parameters also depend on $\tau$, therefore it is necessary to write equations on them, because technological process is non-isothermal:
\begin{linenomath}
\begin{equation}\label{eq_temper}
\dfrac{dT}{d\tau} = \dfrac{\sum_{i=1}^I\left( \dfrac{dx_i}{d\tau} \cdot \Delta H^i_{f,T} \right) }{\sum_i^J \left(x_i C^T_{p_i} \right) },
\end{equation}\label{eq_flow}
\begin{equation}
\dfrac{dF}{d\tau} = \sum_{i=1}^I  \dfrac{x_i}{d\tau}
\end{equation}
\end{linenomath}
with initial condition: at $\tau = 0$, $T(0) = T^0$, where $T$ is temperature, K, $\Delta H^i_{f,T}$ is enthalpy of the $i$-th component formation at temperature $T$, J/mol, $C^T_{p_i}$ is specific heat capacity of the $i$-th component at temperature $T$, (J/mol$\cdot$K), $F$ is mole discharge of the flow, kmol/h.

>>>>

%=================================================================
\documentclass[mathematics,article,accept,pdftex,moreauthors]{Definitions/mdpi} 
% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal and change "submit" to "accept". The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

@@

%=================================================================
% MDPI internal commands
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2022}
\copyrightyear{2022}
\externaleditor{Academic Editor: {Leonid B. Sokolinsky, Mikhail Zymbler}}

\datereceived{31 August 2022} 
%\daterevised{} % Only for the journal Acoustics
\dateaccepted{4 October 2022 } 
\datepublished{} 
%\datecorrected{} % Corrected papers include a "Corrected: XXX" date in the original paper.
%\dateretracted{} % Corrected papers include a "Retracted: XXX" date in the original paper.
\hreflink{https://doi.org/} % If needed use \linebreak
%\doinum{}
%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

@@

%=================================================================
% Full title of the paper (Capitalized)
\Title{On Solving the Problem of Finding Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction Using~a Parallel Global Search Algorithm}

% На английском: On Solving the Problem of Finding Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction Using Parallel Global Search Algorithm

% MDPI internal command: Title for citation in the left column
\TitleCitation{On Solving the Problem of Finding Kinetic Parameters of Catalytic Isomerization of the Pentane-Hexane Fraction Using a Parallel Global Search Algorithm}

@@

% Authors, for the paper (add full first names)
\Author{\hl{Konstantin Barkalov} %MDPI: 1. Please carefully check the accuracy of names and affiliations. 2. Please do not delete our comments. 3. Please revise all questions that we proposed. And reply to them after the corresponding comment. Such as: “Author reply: It should be italic”; “Author reply: I confirm”; “Author reply: I have checked and revised all.” etc..
 $^{1}$\orcidA{}, Irek Gubaydullin $^{2,3}$, Evgeny Kozinov $^{1}$\orcidB{}, Ilya Lebedev $^{1}$\orcidC{}, Roza Faskhutdinova\orcidD{} $^{2,3}$, Azamat Faskhutdinov $^{2}$ and Leniza Enikeeva $^{3,4,}$*\orcidE{}}

@@

\address{%
	$^{1}$ \quad \hl{Lobachevsky} %MDPI: Please add the department/school/faculty/campus in affs. 1, 3 and 4.
 State University of Nizhny Novgorod, \hl{Nizhny Novgorod,} %MDPI: Please add the postal code in affs. 1--4 (or ZIP code in the U.S.). If the postal code is not available, please provide the P.O. Box.
 Russia; konstantin.barkalov@itmm.unn.ru (K.B.); evgeny.kozinov@itmm.unn.ru (E.K.); ilya.lebedev@itmm.unn.ru~(I.L.)\\
	$^{2}$ \quad \hl{Institute} %MDPI: If the information provided presents more than one address, please separate the addresses into different affiliations.
 of Petrochemistry and Catalysis---Subdivision of the Ufa Federal Research Centre of RAS, \mbox{\hl{Ufa}, Russia;} irekmars@mail.ru (I.G.); roza\_fask@mail.ru (R.F.); azamat\_d1989@mail.ru (A.F.)\\
	$^{3}$ \quad Ufa State Petroleum Technological \hl{University}, \hl{Ufa}, Russia\\
	$^{4}$ \quad Novosibirsk State \hl{University}, \hl{Novosibirsk}, Russia
}

% Contact information of the corresponding author
\corres{Correspondence: leniza.enikeeva@yandex.ru}

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{This article is devoted to the problem of developing a kinetic model of a complex chemical reaction using a parallel optimization method. The design of the kinetic model consists of finding the kinetic parameters of the reaction, which cannot be calculated analytically, and since the chemical reaction involves many stages, the optimization problem is {multiextremal}. As a chemical reaction, the process of catalytic isomerization of the pentane-hexane fraction is considered, which is now important due to the switch of the oil refining industry to the production of gasoline corresponding to the Euro-5 standard. {On the basis of known industrial data on the concentrations of reaction components and the temperature at the outlet of the third reactor, the activation energies and pre-exponential factors of each reaction stage were calculated.} To solve the optimization problem, the authors developed a parallel global search algorithm and a program based on Lipschitz optimization. The kinetic parameters found made it possible to develop a mathematical model of the process, which is in good agreement with industrial data. {The developed mathematical model in future works will make it possible to study the dynamics of the gas--liquid flow in the reactor unit, taking into account diffusion and heat exchange processes through the catalyst layer.}}

% Keywords
\keyword{inverse problems of chemical kinetics; mathematical modeling; catalytic isomerisation; Lipschitz optimization; black-box functions; parallel computing}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
\MSC{90C26; 65Y05}
%\JEL{}

@@

	Due to the transition of the domestic oil refining industry to the production of gasoline conforming to the Euro-5 standard, reducing the content of aromatic hydrocarbons is an urgent task, in particular benzene in motor fuel, while maintaining the octane value. In connection with the transition of the oil refining industry to the production of motor gasoline conforming to Euro-5 and Euro-6 standards, the reduction of aromatic hydrocarbons in motor fuel and, mainly, benzene, becomes an urgent task. At the same time, when removing aromatic hydrocarbons, which are high-octane components, it is necessary to preserve the octane number of gasoline. It is known that hydrocarbons with a branched structure have a higher octane number than hydrocarbons with a linear structure. For example, in $n$-pentane, the octane number according to the research method is 62 points and in $iso$-pentane it is 93 points. On an industrial scale, catalytic isomerization units of light paraffins have been introduced to produce hydrocarbons of an isomeric structure. The catalytic isomerization of light paraffins, in turn, makes it possible to obtain a high-octane component of automobile gasoline with a minimum content of aromatic hydrocarbons. The main advantage of this process is the production of high-octane gasoline, while aromatic hydrocarbons are practically not formed. Straight-run gasoline fractions with boiling limits of 62--70 °C containing mainly $n$-pentane and $n$-hexane, pentane-hexane fractions from gas condensate or from gas fractionation plants, as well as catalytic reforming raffinates, are used as feedstock for catalytic isomerization. Modern bifunctional catalysts are used to carry out the process. In order to prevent coke deposition, the catalytic isomerization process is carried out in a hydrogen-containing gas environment under pressure \cite{Hartmut, Fahim, Meyers}.

	Computer information technologies are widely used in the modeling of production processes and today it is difficult to imagine research activities without the use of modern means of constructing and using models. Moreover, special attention is currently being paid to the development of a detailed kinetic model of complex industrial processes \cite{DAO2022172, SAJID20207430, LI2021175, HOU201759}. Based on these models, it is possible to significantly increase the efficiency of the process while utilizing minimal energy and material resources \cite{uskovfibrous, en13133393}. The main advantage of such methods is the absence of additional costs for the construction of pilot plants. The developed model should describe the process within a wide range of operating conditions, have high flexibility and allow for changes and calculations of the composition of the reaction mixture, etc.

	{The reactor unit of the catalytic isomerization unit of the pentane-hexane fraction is the object of research in this work. The production process was carried out in a continuous way. The reactor unit consisted of three consecutive reactors.
		The raw material was a light gasoline fraction after hydrotreating. The catalyst of the process was a bifunctional catalyst of the SI-2 brand. The total mass of the catalyst in the reactor unit was 27 tons (9 tons in each reactor). The process was carried out at a constant pressure of 3.2 MPa.
		For calculations, data for three days of operation of an industrial installation were taken.
		Based on chromatographic analyses and mass flow rates for each day, the component compositions of the reaction mixture were formed both at the entrance to the reactor unit and at the exit from the reactor~unit.}

	In \cite{enik2021} we already reviewed this process, but in order to simplify the model, the scheme of chemical transformations, in particular hydrocracking reactions, have been revised. Thus, the most probable hydrocracking reactions, based on the carbonium-ion mechanism, were formed. The new transformation scheme implies the recalculation of the kinetic parameters of the reaction, which form a mathematical model of the process.

	In \cite{enik2021}, a genetic algorithm was used to calculate kinetic parameters, and the calculation time of kinetic parameters was about 10 h. Therefore, in order to develop a mathematical model of a new transformation scheme, the question of choosing a more effective optimization method arises.

	In the complex problems of chemical kinetics, kinetic constants of reactions cannot be found analytically; they can only be found using numerical methods (see, for example,~\cite{Enikeeva2020}). In this case, the quality criterion (objective function) is set in the form of a certain algorithm, which includes the stage of numerical modeling, which, in turn, makes it computationally intensive. Furthermore, in the inverse problems of chemical kinetics, the objective function can be substantially multiextremal, i.e., having many local extremes along the global one. 
	{Gradient-based optimization methods will not provide a good result for such problems. On the one hand, such methods converge only to a local solution. On the other hand, in the considered problem, the gradient of the objective function is not known analytically and its numerical estimate will be a time-consuming operation.}

	Computational algorithms for solving such multiextremal problems (global optimization methods) differ vastly from local optimization techniques since they require the study of the entire search domain (see, for example, \cite{PaulaviciusZilinskas2014,Sergeyev2017}). These methods can be divided into two classes: metaheuristic and deterministic. Metaheuristic algorithms are usually based on simulations of processes occurring in the wild.
	Examples of metaheuristic algorithms are simulated annealing and evolutionary computing, etc. (see, for example, \cite{Battiti2009,Eiben2015}).
	These algorithms are included in many software libraries for solving global optimization problems (for example, MATLAB Global Optimization Toolbox, \hl{Scipy.Optimize,} %MDPI: Please check if there should be a space after ``Scipy.''.
 etc.).
	Due to the operation speed and low requirements for computing resources (in particular, memory), metaheuristic algorithms have become widespread for solving practical problems. However, the solution to the problem found by such an algorithm is, in general, local, and can be located far from the global optimum \cite{Kvasov2018}.

	It is also possible to construct deterministic methods of global search, which are different from brute force algorithms and metaheuristic algorithms. Such methods are associated with the presence and consideration of a priori assumptions about the problem properties. 
	One of the natural assumptions about the problem is that changing the values of an objective function is limited by changes of its argument. In this case, the function satisfies the Lipschitz condition, and the problem itself is called the Lipschitz global optimization problem. The central direction of the development of modern methods of Lipschitz optimization is the development of parallel algorithms. It is assumed that the values of the objective function are determined as a result of solving some other task and require a costly computational experiment.
	{On many benchmarks, Lipschitz global optimization methods outperform metaheuristic algorithms in terms of the number of correctly solved problems~\cite{Sergeyev2018,Sovrasov2019}.}

	This paper reflects the results of applying parallel methods of Lipschitz optimization developed in \cite{Barkalov2016,Strongin2018,Sysoyev2017} to the solution of inverse problems of chemical kinetics.  
	The main part of the paper is structured as follows.

	Section \ref{sec_matmodel} provides a description of the mathematical model of the chemical reaction under study. Section \ref{sec_GSA} discusses the standard formulation of the Lipschitz global optimization problem and the algorithm for solving it. Section \ref{sec_numexper} outlines the issues related to the technical aspects of the implementation of the parallel algorithm. The physico-chemical interpretation of the results of numerical calculations using the developed parallel algorithm is presented in Section \ref{sec_discussion}. 
	%The last section contains the conclusion and plans for future work.

	\section{Mathematical Model}\label{sec_matmodel}

	Let us consider the mechanism of chemical transformations of the isomerization of the pentane-hexane fraction. The reaction consists of 48 reaction stages, which are presented in Table \ref{tab_chemistry}. The following abbreviations were adopted: 2-MP is 2-methylpentane; 3-MP is 3-methylpentane; 2,2-DMB is 2,2-dimethylbutane; 2,3-DMB is 2,3-dimethylbutane; CH* is cyclohexane; MCP is methyl cyclopentane; B is benzene; CP is cyclopentane.

	The mathematical model of the chemical process is a non-linear system of differential equations \cite{enik2021,logan}:
	\begin{linenomath}
		\begin{equation}\label{eq_concentrations}
			\dfrac{dx_i}{d\tau} = \sum_{j=1}^J \nu_{ij} w_j, i = 1,\dots,I,
		\end{equation}
		\begin{equation}\label{eq_rate}
			w_{j} = k_j \prod_{i=1}^I \left( \dfrac{x_i}{F} \right),
		\end{equation}
		\begin{equation}\label{eq_constant}
			k_j = k^0_j \exp \left( -\dfrac{E_j}{RT} \right),
		\end{equation}
	\end{linenomath}
	with initial values at $\tau = 0$, $x_i(0) = x_i^0$, where $\nu_{ij}$ are stoichiometric coefficients of stages of the chemical reaction; $J$ is the number of stages ($J = 48$), $x_i$ is the molar flow rate of the $i$-th component of reaction, kmol/h; $I$ is the number of components ($I = 17$); $w_j$ is the rate of the $j$-th stage of reaction, kmol/(h$\cdot$kg cat.), $E_j$ is the activation energy of the reaction stage, J/mol; $R$ is the universal gas constant ($R$ = 8.31 J/(mol$\cdot$K), $T$ is the temperature, K; $k_j$ are pre-exponential factors, kmol/(h$\cdot$kg cat.); $\tau$ is conditioned contact time, (kg cat.); $F$ is the total molar flow rate, kmol/h.

	The mathematical model involves parameters such as the raw material flow rate $F$ in the reactor and temperature $T$, but these two parameters also depend on $\tau$; therefore it is necessary to write equations on them because the technological process is non-isothermal:
	\begin{linenomath}
		\begin{equation}\label{eq_temper}
			\dfrac{dT}{d\tau} = \dfrac{\sum_{i=1}^I\left( \dfrac{dx_i}{d\tau} \cdot \Delta H^i_{f,T} \right) }{\sum_i^J \left(x_i C^T_{p_i} \right) },
		\end{equation}
		\begin{equation}\label{eq_flow}
			\dfrac{dF}{d\tau} = \sum_{i=1}^I  \dfrac{x_i}{d\tau},
		\end{equation}
	\end{linenomath}
	with initial condition: at $\tau = 0$, $T(0) = T^0$, where $T$ is the temperature, K, $\Delta H^i_{f,T}$ is the enthalpy of the $i$-th component formation at temperature $T$, J/mol, $C^T_{p_i}$ is the specific heat capacity of the $i$-th component at temperature $T$, (J/mol$\cdot$K), $F$ is the mole discharge of the flow, kmol/h.
